Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-03 paper/paper.qmd
Include abstract/importance in github markdown
Use Quarto conditional content to only include them
if the format is markdown and they are defined in
the yaml header.

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
bibliography: references.bib
fontsize: 11pt
date: last-modified
date-format: medium
prefer-html: true
format:
  manuscript-pdf:
    keep-tex: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
#abstract: TODO
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
---
```{r setup, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data, message = FALSE}
library(assertthat)
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)

load(here("results", "stats.RData"))
table_font_size <- 10
```

@@

prevalence of cdi. prevalence of severe cdi outcomes.
antibiotics typical risk factor for cdi, but nonantibiotic medications can 
increase susceptibilty too [@tomkovich_osmotic_2021]
and non-nosomial CDI on the rise

@@

prediction models based on EHR for whether infection occurs in the first place already in use.
so how about predicting severity of infections to guide treatment?
also models on EHR to predict adverse outcomes (Li, Rao).
OTUs vs EHRs to predict severity.
CDI severity prediction models could be deployed to screen patients at risk and
guide clinicians to consider prescribing a different course of treatment.
When paired with treatment options that may reduce risk of severity, deploying
prediction models can guide clinician decision-making to improve patient outcomes
while minimizing unnecessary harms.

@@

After training on the full datasets, the performance as measured by the AUROCs
of the training set cross-validation folds were similar to those of the held-out
test sets, indicating that the models are neither overfit nor underfit
(@fig-performance A).
As measured by AUROC on the held-out test sets, models predicting pragmatic
severity performed best with a median AUROC of `r perf_medians$auroc_full_pragmatic`,
and this was significantly different from that of the other definitions 
on the full datasets (P < 0.05).
Models predicting IDSA, all-cause, and attributable severity performed similarly
with median test set AUROCs of
`r perf_medians$auroc_full_idsa`,
`r perf_medians$auroc_full_allcause`, and
`r perf_medians$auroc_full_attrib` respectively.
The test set AUROCs were not significantly different (P > 0.05) for attributable
and IDSA nor for attributable and all-cause, but the IDSA and all-cause AUROCs
were significantly different from each other (P < 0.05).
We plotted the receiver-operator characteristic curve and found that the
pragmatic severity models outperformed the others at all specificity values
(@fig-performance B).
A prior study trained a logistic regression model on whole Electronic Health
Record data extracted on the day of CDI diagnosis to predict attributable
severity, yielding an AUROC of `r ehr_auroc` [@li_using_2019].
While our OTU-based attributable severity model did not meet this performance,
the OTU-based pragmatic severity model performed just as well as the EHR-based
model in terms of AUROC.

@@

We performed permutation feature importance to determine which OTUs contributed
the most to model performance.
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits, with greater differences in AUROC meaning
greater importance.
We plotted mean decrease in AUROC alongside log~10~-transformed mean
relative abundances for the top OTUs (@fig-features).
There is not always a clear pattern of increased or decreased relative abundance
for important OTUs in severe CDI cases, but all of the top 5 OTUs had an
increased mean relative abundance in severe cases relative to not severe cases.
_Enterococcus_ was the most important OTU, being significantly important for 
all models except for attributable severity on the full dataset.
Enrichment of _Enterococcus_ in _C. difficile_ infection and severity has been well-documented
in prior studies, thus its importance and increase in abundance for severe cases
is not suprising 
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
_Staphylococcus_ was important for the pragmatic and all-cause definitions on
the full datasets, but not for models trained on the intersection dataset.
_Lactobacillus_ was important only for the all-cause definition on the intersection
dataset, and its enrichment during _C. difficile_ infection has also been shown
in prior studies 
[@antharam_intestinal_2013;
@berkell_microbiota-based_2021].
All remaining OTUs had differences in AUROC < 0.02 and were only significantly
important in one or two of the models at most.
For many of the top OTUs, there is wide variance in importance, perhaps due to
the imbalanced nature of the severity outcomes.

@@

Even if a model performs well, it may not be useful in a clinical setting unless
it can guide clinicians to choose between treatment options.
At this time, we are not aware of any direct evidence that a particular
treatment reduces the risk of severe CDI outcomes.
However, with some assumptions we offer a proof-of-concept analysis of the
potential clinical value of OTU-based severity prediction models when paired
with treatments that may reduce severity.
When considering the suitability of a model for deployment in clinical settings,
the number needed to screen (NNS) is a highly relevant metric representing how
many patients must be predicted as severe by the model to identify one true
positive.
Similarly, the number needed to treat (NNT) is the number of true positive
patients that must be treated by an intervention in order for one patient to
benefit from the treatment.
Multiplying NNS by NNT yields the number needed to benefit (NNB): the number
of patients predicted to have a severe outcome who then benefit from the
treatment [@liu_number_2019].
Thus the NNB pairs model performance with treatment effectiveness to estimate
the benefit of using predictive models in clinical practice.

@@

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model
(@tbl-risk).
We excluded the IDSA severity models as the IDSA severity scores were calculated
on the day of diagnosis, thus they are classification rather than prediction
problems.
Furthermore, IDSA severity scores do not correlate well with disease-related
adverse events which are a more salient outcome to prevent.
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data extracted two days after diagnosis and from a
smaller set of manually curated variables, achieving precision values of 0.417
(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
curated model at the 95th percentile of risk
[@li_using_2019;@rao_clostridium_2015]. <!-- TODO what about NNS on day of diagnosis? -->
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin hold
true, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model in order for one patient to benefit.
For comparison, pairing the prior EHR-based model with fidaxomicin would yield
an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
one patient to benefit.
These estimates represent a proof-of-concept demonstration of the potential
value of deploying severity prediction models to guide clinicians' treatment
decisions.

<!--
rough estimate of costs.
current: everyone gets vancomycin.
potential: patients flagged as severe get fidaxomicin. based on NNB, estimate
how much money saved in averting severe outcomes.
-->

@@

Discuss important OTUs. which ones concord with literature, which ones may be new.
Abundance data are sparse, likely due to these patients being on antibiotics.
Really showcases importance of having as many samples as possible when data are
sparse and the outcome is low prevalence.
we do not know which antibiotics were prescribed to treat these CDI cases, nor
which antibiotics patients may have taken prior to the CDI.
differences in microbiota between patients may be due to different abx...
different antibiotics have been shown to create different forms of dysbiotic
microbiota [@berkell_microbiota-based_2021] and
differential cdi clearance [@tomkovick_initial_2020].
vanc-resistant enterococcus new nosocomial alliance [@poduval_clostridium_2000].

Compare to EHR-based models.

@@

This study was approved by the University of Michigan Institutional Review
Board. All patient samples were collected by the University of Michigan Health
System from January 2016 through December 2017.
Stool samples that had unformed stool consistency were tested for *C. difficile* by
the clinical microbiology lab with a two-step algorithm that included detection
of *C. difficile* glutamate dehydrogenase and toxins A and B by enzyme
immunoassay with reflex to PCR for the *tcdB* gene when results were discordant.
1,517 stool samples were collected from patients diagnosed with a CDI.
Leftover stool samples that were sent to the clinical microbiology lab were
collected and split into different aliquots.
For 16S sequencing, the aliquot of stool was re-suspended in
DNA genotek stabilization buffer and then stored in the -80&deg;C freezer.
Only the first CDI sample per patient was used for subsequent ML analyses such
that no patient is represented more than once, resulting in a dataset of
`r n_cases_first` samples.

@@

All sequences were processed with mothur (v1.46) using the MiSeq SOP protocol
[@schloss_introducing_2009; @kozich_development_2013].
Paired sequencing reads were combined and
aligned with the SILVA (v132) reference database [@quast_silva_2013] and taxonomy was
assigned with a modified version of the Ribosomal Database Project reference
sequences (v16) [@cole_ribosomal_2014].
Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
mothur [@westcott_opticlust_2017], resulting in `r num_otus` OTUs.
<!-- TODO supplementary figure with alpha and beta diversity & significance.
Samples were rarefied to 5,000 sequences per sample, repeated 1,000
times for alpha and beta diversity analysis.
 -->

@@

::: {#fig-flowchart fig-cap="**CDI severity definitions.**
**A)** Decision flow chart to define CDI cases as severe according to the
Infectious Diseases Society of America (IDSA) based on lab values,
the occurrence of an adverse outcome due to any cause (All-cause),
and the occurrence of disease-related complications confirmed as attributable to
CDI with chart review (Attributable).
**B)** The proportion of severe CDI cases labelled according to each definition.
An additional 'Pragmatic' severity definition uses the Attributable definition
when possible, and falls back to the All-cause definition when chart review is
not available.
<!-- TODO table (supplementary?) showing counts & frequency of positives-->
"}
![](figures/flowchart_sankey.png)
:::

>>>>

---
title: "Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome"
date: last-modified
date-format: medium
prefer-html: true
bibliography: references.bib
fontsize: 11pt
format:
  manuscript-pdf:
    keep-tex: true
    linenumbers: true
    doublespacing: true
    lof: false
  html:
    embed-resources: true
runningtitle: Predicting CDI Severity from OTUs
runningauthor: Sovacool
papertype: Preprint
author:
  - name: Kelly L. Sovacool
    orcid: 0000-0003-3283-829X
    affiliations:
      - ref: bioinf
  - name: Sarah E. Tomkovich
    affiliations:
      - ref: micro
  - name: Megan L. Coden
    affiliations:
      - ref: mcdb
  - name: Vincent B. Young
    affiliations:
      - ref: micro
      - ref: intmed
  - name: Krishna Rao
    affiliations:
      - ref: intmed
  - name: Patrick D. Schloss
    orcid: 0000-0002-6935-4275
    email: pschloss@umich.edu
    corresponding: true
    affiliations:
      - ref: micro
      - ref: ccmb
affiliations:
  - id: bioinf
    name: Department of Computational Medicine & Bioinformatics, University of Michigan
  - id: micro
    name: Department of Microbiology & Immunology, University of Michigan
  - id: mcdb
    name: Department of Molecular, Cellular, and Developmental Biology, University of Michigan
  - id: intmed
    name: Division of Infectious Diseases, Department of Internal Medicine, University of Michigan
  - id: ccmb
    name: Center for Computational Medicine and Bioinformatics, University of Michigan
abstract: |
  _Clostridioides difficile_ (_C. difficile_) infection (CDI) can lead to 
  adverse outcomes including ICU admission, colectomy, and death.
  The composition of the gut microbiome plays an important role in determining
  colonization resistance and clearance upon exposure to _C. difficile_.
  We investigated whether machine learning (ML) models trained on gut
  microbiome compositions could predict which CDI cases led to severe outcomes.
  We collected 1,277 stool samples from CDI patients on the day of diagnosis and
  characterized the gut microbiome via 16S rRNA gene amplicon sequencing.
  We processed and clustered the sequences into _de novo_ Operational Taxonomic Units
  (OTUs), which we used to train ML models for predicting whether a severe outcome
  occurred according to three different severity definitions.
  We learned that our OTU-based Random Forest models perform best when predicting
  attributable severity, defined as an adverse outcome within 30 days of CDI
  diagnosis confirmed as attributable to CDI following chart review by a clinician.
  This finding suggests that chart review is valuable to verify the cause of
  severe outcomes.
  We used permutation importance to identify which OTUs contributed most to model
  performance and found that _Enterococcus_, which has previously been identified as
  positively correlated with _C. difficile_ colonization, was the largest contributor.
  In all, our results show that ML models can modestly identify patients at risk
  for CDI, which will allow healthcare providers to personalize treatment options
  to protect patients likely at risk and ultimately improve CDI outcomes.
#importance: TODO
keywords: _C. difficile_ infection, supervised machine learning, gut microbiome, amplicion sequencing
execute:
  eval: true
  echo: false
  output: false
  cache: true
  message: false
  warning: false
---
```{r setup}
schtools::set_knitr_opts()
knitr::opts_chunk$set(dpi = 600)
```
```{r load_data}
library(assertthat)
library(glue)
library(here)
library(kableExtra)
library(knitr)
library(tidyverse)

load(here("results", "stats.RData"))
table_font_size <- 10
assert_that(n_cases_first == 1277)
```

::: {.content-visible when-format="markdown"}

```{r abstract}
#| output: asis
if ('abstract' %in% names(rmarkdown::metadata)) {
  cat('# Abstract\n', rmarkdown::metadata$abstract)
}
if ('importance' %in% names(rmarkdown::metadata)) {
  cat('# Importance\n', rmarkdown::metadata$importance)
}
```
:::

@@

prevalence of cdi. prevalence of severe cdi outcomes.
antibiotics typical risk factor for cdi, but nonantibiotic medications can 
increase susceptibilty too [@tomkovich_osmotic_2021]
and non-nosocomial CDI on the rise

@@

prediction models based on EHR for whether infection occurs in the first place already in use [@otles_clostridioides_2023].
so how about predicting severity of infections to guide treatment.
also models on EHR to predict adverse outcomes (Li, Rao).
serum-based biomarker model (Dieterle mbio 2020).
OTUs vs EHRs to predict severity.
CDI severity prediction models could be deployed to screen patients at risk and
guide clinicians to consider prescribing a different course of treatment.
When paired with treatment options that may reduce risk of severity, deploying
prediction models can guide clinician decision-making to improve patient outcomes
while minimizing unnecessary harms.

@@

After training on the full datasets, the performance as measured by the AUROCs
of the training set cross-validation folds were similar to those of the held-out
test sets, indicating that the models are neither overfit nor underfit
(@fig-performance A).
As measured by AUROC on the held-out test sets, models predicting pragmatic
severity performed best with a median AUROC of `r perf_medians$auroc_full_pragmatic`,
and this was significantly different from that of the other definitions 
on the full datasets (P < 0.05).
Models predicting IDSA, all-cause, and attributable severity performed similarly
with median test set AUROCs of
`r perf_medians$auroc_full_idsa`,
`r perf_medians$auroc_full_allcause`, and
`r perf_medians$auroc_full_attrib` respectively.
The test set AUROCs were not significantly different (P > 0.05) for attributable
and IDSA nor for attributable and all-cause, but the IDSA and all-cause AUROCs
were significantly different from each other (P < 0.05).
We plotted the receiver-operator characteristic curve and found that the
pragmatic severity models outperformed the others at all specificity values
(@fig-performance B).
For comparison, a prior study trained a logistic regression model on whole
Electronic Health Record data extracted on the day of CDI diagnosis to predict
attributable severity, yielding an AUROC of `r ehr_auroc` [@li_using_2019].
While our OTU-based attributable severity model did not meet this performance,
the OTU-based pragmatic severity model performed just as well as the EHR-based
model in terms of AUROC.

@@

We performed permutation feature importance to determine which OTUs contributed
the most to model performance.
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits, with greater differences in AUROC meaning
greater importance.
We plotted mean decrease in AUROC alongside log~10~-transformed mean
relative abundances for the top OTUs (@fig-features).
There is not always a clear pattern of increased or decreased relative abundance
for important OTUs in severe CDI cases, but all of the top 5 OTUs had an
increased mean relative abundance in severe cases relative to not severe cases.
_Enterococcus_ was the most important OTU, being significantly important for 
all models except for attributable severity on the full dataset.
_Staphylococcus_ was important for the pragmatic and all-cause definitions on
the full datasets, but not for models trained on the intersection dataset.
_Lactobacillus_ was important only for the all-cause definition on the intersection
dataset.
All remaining OTUs had differences in AUROC < 0.02 and were only significantly
important in one or two of the models at most.
TODO summarize which OTUs enriched in severe vs not severe.
TODO concluding sentence.

@@

Even if a model performs well, it may not be useful in a clinical setting unless
it can guide clinicians to choose between treatment options.
At this time, we are not aware of any direct evidence that a particular
treatment reduces the risk of severe CDI outcomes.
However, with some assumptions we offer a proof-of-concept analysis of the
potential clinical value of OTU-based severity prediction models when paired
with treatments that may reduce severity.
When considering the suitability of a model for deployment in clinical settings,
the number needed to screen (NNS) is a highly relevant metric representing how
many patients must be predicted as severe by the model to identify one true
positive.
Similarly, the number needed to treat (NNT) is the number of true positive
patients that must be treated by an intervention in order for one patient to
benefit from the treatment.
Multiplying NNS by NNT yields the number needed to benefit (NNB): the number
of patients predicted to have a severe outcome who then benefit from the
treatment [@liu_number_2019].
Thus the NNB pairs model performance with treatment effectiveness to estimate
the benefit of using predictive models in clinical practice, and are useful for
comparing models and performing cost-benefit analyses.

@@

To calculate a clinically-relevant NNS for these models, we computed the
confusion matrix at the 95th percentile of risk for each prediction model
(@tbl-risk).
We excluded the IDSA severity models as the IDSA severity scores were calculated
on the day of diagnosis, thus they are classification rather than prediction
problems.
Furthermore, IDSA severity scores do not correlate well with disease-related
adverse events which are a more salient outcome to prevent.
Among the models predicting severe outcomes, those trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
Multiplying the NNS of the OTU-based models by the estimated NNT of 
`r fdx_nnt` for fidaxomicin yields NNB values of `r allcause_fdx_nnb` 
for all-cause severity, `r attrib_fdx_nnb` for attributable severity,
and `r pragmatic_fdx_nnb` for pragmatic severity.
Thus, in a hypothetical scenario where these assumptions about fidaxomicin hold
true, between `r min_fdx_nnb` and `r max_fdx_nnb` patients would need to be
predicted to experience a severe outcome and be treated with fidaxomicin in
order for one patient to benefit.
As the NNS values were computed at the 95th percentile of risk (where 5% of
patients screened are predicted to experience severity), these NNB values mean
that `r min_screen` to `r max_screen` total CDI patients would need to be
screened by an OTU-based prediction model in order for one patient to benefit.
For comparison, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data extracted two days after diagnosis and from a
smaller set of manually curated variables, achieving precision values of 0.417
(NNS = `r ehr_nns`) for the EHR model and 0.167 (NNS = `r curated_nns`) for the
curated model at the 95th percentile of risk
[@li_using_2019;@rao_clostridium_2015]. <!-- TODO possible to find NNS on day of diagnosis for EHR model-->
Pairing the prior EHR-based model with fidaxomicin would yield
an NNB of `r ehr_fdx_nnb` with `r ehr_screen` total CDI patients screened for
one patient to benefit, although the EHR was extracted two days after diagnosis
while OTUs in this study are from stool samples collected on the day of diagnosis.
These estimates represent a proof-of-concept demonstration of the potential
value and trade-offs of deploying severity prediction models trained on
microbial factors versus EHRs to guide clinicians' treatment decisions.

<!--
rough estimate of treatment costs. exclude others like cost of 16S sequencing & deployment.
current: everyone gets vancomycin.
potential: patients flagged as severe get fidaxomicin. based on NNB, estimate
how much money saved in averting severe outcomes.
-->

@@

Discuss important OTUs. which ones concord with literature, which ones may be new.
For many of the top OTUs, there is wide variance in importance, perhaps due to
the imbalanced nature of the severity outcomes.
Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and severity has been well-documented
in prior studies, thus its importance and increase in abundance for severe cases
is not surprising 
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
Abundance data are sparse, likely due to these patients being on antibiotics.
Really showcases importance of having as many samples as possible when data are
sparse and the outcome is low prevalence.
we do not know which antibiotics were prescribed to treat these CDI cases, nor
which antibiotics patients may have taken prior to the CDI.
differences in microbiota between patients may be due to different abx...
different antibiotics have been shown to create different forms of dysbiotic
microbiota [@berkell_microbiota-based_2021] and
differential cdi clearance [@tomkovich_initial_2020].
vanc-resistant enterococcus new nosocomial alliance [@poduval_clostridium_2000].

Compare to EHR-based models.
in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be.
further work needed to characterize the costs and benefits.

@@

This study was approved by the University of Michigan Institutional Review
Board. All patient samples were collected by the University of Michigan Health
System from January 2016 through December 2017.
Stool samples that had unformed stool consistency were tested for *C. difficile* by
the clinical microbiology lab with a two-step algorithm that included detection
of *C. difficile* glutamate dehydrogenase and toxins A and B by enzyme
immunoassay with reflex to PCR for the *tcdB* gene when results were discordant.
1,517 stool samples were collected from patients diagnosed with a CDI.
Leftover stool samples that were sent to the clinical microbiology lab were
collected and split into different aliquots.
For 16S sequencing, the aliquot of stool was re-suspended in
DNA genotek stabilization buffer and then stored in the -80&deg;C freezer.

@@

All sequences were processed with mothur (v1.46) using the MiSeq SOP protocol
[@schloss_introducing_2009; @kozich_development_2013].
Paired sequencing reads were combined and
aligned with the SILVA (v132) reference database [@quast_silva_2013] and taxonomy was
assigned with a modified version of the Ribosomal Database Project reference
sequences (v16) [@cole_ribosomal_2014].
Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
mothur [@westcott_opticlust_2017], resulting in `r num_otus` OTUs.
Only the first CDI sample per patient was used for subsequent ML analyses such
that no patient is represented more than once, resulting in a dataset of
`r n_cases_first` samples.
<!-- TODO supplementary figure with alpha and beta diversity & significance.
Samples were rarefied to 5,000 sequences per sample, repeated 1,000
times for alpha and beta diversity analysis.
 -->

@@

::: {#fig-flowchart fig-cap="**CDI severity definitions.**
**A)** Decision flow chart to define CDI cases as severe according to the
Infectious Diseases Society of America (IDSA) based on lab values,
the occurrence of an adverse outcome due to any cause (All-cause),
and the occurrence of disease-related complications confirmed as attributable to
CDI with chart review (Attributable).
**B)** The proportion of severe CDI cases labelled according to each definition.
An additional 'Pragmatic' severity definition uses the Attributable definition
when possible, and falls back to the All-cause definition when chart review is
not available.
See @tbl-counts for sample counts and proportions of severe cases across
severity definitions.
"}
![](figures/flowchart_sankey.png)
:::
